---
output:
  pdf_document: default
  html_document: default
---

# Instrumental Variable 2: Implementation in R

## Example 1: Wage regression

- Use dataset "Mroz", cross-sectional labor force participation data that accompany "Introductory Econometrics" by Wooldridge. 
    - Original data from  *"The Sensitivity of an Empirical Model of Married Women's Hours of Work to Economic and Statistical Assumptions"* by Thomas Mroz published in *Econometrica* in 1987.
    - Detailed description of data: https://www.rdocumentation.org/packages/npsf/versions/0.4.2/topics/mroz 
    
 
```{r}

library("foreign")

# You might get a message "cannot read factor labels from Stata 5 files", but you do not have to worry about it. 
data <- read.dta("MROZ.DTA")


```

- Describe data
```{r}
library(stargazer)
stargazer(data, type = "text")
```

- Consider the wage regression 
$$
\log(w_i) = \beta_0 + \beta_1 educ_i + \beta_2 exper_i + \beta_3 exper_i^2 + \epsilon_i
$$
- We assume that $exper_i$ is exogenous but $educ_i$ is endogenous.
- As an instrument for $educ_i$, we use the years of schooling for his or her father and mother, which we call $fathereduc_i$ and $mothereduc_i$. 
- Discussion on these IVs will be later. 

```{r}

library("AER")
library("dplyr")

# data cleaning
data %>% 
  select(lwage, educ, exper, expersq, motheduc, fatheduc) %>%
  filter( is.na(lwage) == 0 ) -> data

# OLS regression
result_OLS <- lm( lwage ~ educ + exper + expersq, data = data)

# IV regression using fathereduc and mothereduc
result_IV <- ivreg(lwage ~ educ + exper + expersq | fatheduc + motheduc + exper + expersq, data = data)

# Robust standard errors 
# gather robust standard errors in a list
rob_se <- list(sqrt(diag(vcovHC(result_OLS, type = "HC1"))),
               sqrt(diag(vcovHC(result_IV, type = "HC1"))))

# Show result
stargazer(result_OLS, result_IV, type ="text", se = rob_se)


```

- How about the first stage? You should always check this!!

```{r}

# First stage regression 

result_1st <- lm(educ ~ motheduc + fatheduc + exper + expersq, data = data)

# F test
linearHypothesis(result_1st, 
                 c("fatheduc = 0", "motheduc = 0" ), 
                 vcov = vcovHC, type = "HC1")

```

### Discussion on IV

- Labor economists have used family background variables as IVs for education. 
- Relevance: OK from the first stage regression. 
- Independence: A bit suspicious. Parents' education would be correlated with child's ability through quality of nurturing at an early age. 
- Still, we can see that these IVs can mitigate (though may not eliminate completely) the omitted variable bias. 
- Discussion on the validity of instruments is crucial in empirical research. 


