---
output:
  pdf_document: default
  html_document: default
---

# Panel Data 2: Implementation in R

## Preliminary:

- I use the following package
    - `lfe` package.

## Panel Data Regression

- I use the dataset `Fatalities` in `AER` package.
    - See https://www.rdocumentation.org/packages/AER/versions/1.2-6/topics/Fatalities for details.

```{r}

library(AER)
data(Fatalities)

str(Fatalities)

```

- As a preliminary analysis, let's plot the relationship between fatality rate and beer tax in 1998.

```{r}
library("dplyr")

Fatalities %>%
  mutate(fatal_rate = fatal / pop * 10000) %>% 
  filter(year == "1988") -> data


plot(x = data$beertax, 
     y = data$fatal_rate, 
     xlab = "Beer tax (in 1988 dollars)",
     ylab = "Fatality rate (fatalities per 10000)",
     main = "Traffic Fatality Rates and Beer Taxes in 1988",
     pch = 20, 
     
     col = "steelblue")



```

- Positive correlation between alcohol tax and traffic accident. Possibly due to omitted variable bias. 

- Run fixed effect regression using `felm` command in `lfe` package. 
    - https://www.rdocumentation.org/packages/lfe/versions/2.8-3/topics/felm
    

```{r}

library("lfe")

Fatalities %>%
    mutate(fatal_rate = fatal / pop * 10000) -> data

# OLS 
result_ols <- felm( fatal_rate ~ beertax  | 0 | 0 | 0, data = data )
summary(result_ols, robust = TRUE)

# State FE
result_stateFE <- felm( fatal_rate ~ beertax  | state | 0 | state, data = data )
summary(result_stateFE, robust = TRUE)

# State and Year FE
result_bothFE <- felm( fatal_rate ~ beertax  | state + year | 0 | state, data = data )
summary(result_bothFE, robust = TRUE)


stargazer::stargazer(result_ols, result_stateFE, result_bothFE, type = "text")


```

- What if we do not use the cluster-robust standard error?

```{r}

# State FE w.o. CRS
result_wo_CRS <- felm( fatal_rate ~ beertax  | state | 0 | 0, data = data )

# State FE w. CRS
result_w_CRS <- felm( fatal_rate ~ beertax  | state | 0 | state, data = data )

# Report heteroskedasticity robust standard error and cluster-robust standard errors
stargazer::stargazer(result_wo_CRS, result_w_CRS,  type = "text", se = list(summary(result_wo_CRS)$rse, NULL))

```


## Panel Data with Instrumental Variables

- Revisit the demand for Cigaretts
- Consider the following model
$$
\log (Q_{it}) = \beta_0 + \beta_1 \log (P_{it}) + \beta_2 \log(income_{it}) + u_i + e_{it}
$$
where 
    - $Q_{it}$ is the number of packs per capita in state $i$ in year $t$, 
    - $P_{it}$ is the after-tax average real price per pack of cigarettes, and 
    - $income_{it}$ is the real income per capita. This is demand shifter.
- As an IV for the price, we use the followings:
    - $SalesTax_{it}$: the proportion of taxes on cigarettes arising from the general sales tax.
        - Relevant as it is included in the after-tax price
        - Exogenous(indepndent) since the sales tax does not influence demand directly, but indirectly through the price.
    - $CigTax_{it}$: the cigarett-specific taxes
    
    
```{r}

# load the data set and get an overview
library(AER)
data("CigarettesSW")
CigarettesSW %>% 
  mutate( rincome = (income / population) / cpi) %>% 
  mutate( rprice  = price / cpi ) %>% 
  mutate( salestax = (taxs - tax) / cpi ) %>% 
  mutate( cigtax = tax/cpi ) -> Cigdata


```
    
- Run IV regression with panel data.

```{r}

# OLS
result_1 <- felm( log(packs) ~ log(rprice) + log(rincome)  | 0 | 0 | state, data = Cigdata )

# State FE
result_2 <- felm( log(packs) ~ log(rprice) + log(rincome)  | state | 0 | state, data = Cigdata )

# IV without FE
result_3 <- felm( log(packs) ~ log(rincome)  | 0 | (log(rprice) ~  salestax + cigtax) | state, data = Cigdata )

# IV with FE 
result_4 <- felm( log(packs) ~ log(rincome)  | state | (log(rprice) ~  salestax + cigtax) | state, data = Cigdata )

stargazer::stargazer(result_1, result_2, result_3, result_4, type = "text")


```


